Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-01-27 resex.tex
finished section about 7th turn pulsing

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total. Both schemes are illustrated in Fig.~\ref{fig:fill}.
\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment
		\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
	\centering
	2017 experiment
	\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom).}
\end{figure*}

@@

	\caption{\label{fig:patternfma} FMA analysis for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H further confirms that the $7Q_x$ resonance is the main resonance driven. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{sec:fma:10}, Fig.~\ref{fig:fma:10}).}

@@

\subsubsection{10th turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude as shown in Figs.~\ref{fig:10thexp} and \ref{fig:10thexpsim}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane shown in Fig.~\ref{fig:10thexp}.
	\item Change of the beam distribution due to the resonant excitation illustrated in Fig.~\ref{fig:10thexp} and \ref{fig:10thexpprof}.
	\item As loss rates, emittance growth and distribution stayed unchanged for the $4$~reference bunches, the above observations can be truly associated with the effect of the resonant excitation.	
\end{enumerate}
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{bunch_intensity_v10th_no_damper_avg.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emith_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emitv_avg_rel_v10th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and emittance (bottom) obtained during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}
\end{figure}

Besides the determination of the scaling of the loss rate and emittance growth with the excitation amplitude as required for the specification of the hollow electron lens, the most novel observation is the direct measurement of the change of the beam distribution, here with the Beam Synchrotron Radiation Telescope (BSRT)~\cite{bsrtprofinj}. As an example for the change in distribution, the vertical profile of a reference bunch and one bunch experiencing the maximum excitation of $5\cdot\Delta A=A_{\mathrm{max}}$ are compared in Fig.~\ref{fig:10thexpprof}. The distribution of the excited bunch clearly changes and becomes non-Gaussian, while the reference bunch stays unchanged.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch, no excitation
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_50.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch, $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_1300.png}
	\end{minipage}
	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. The black line in the plot of the residual is the current profile minus the initial profile and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis see \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}
\end{figure}

Comparing these experimental results with the simulations including only the excitation through the $10^{\mathrm{th}}$~turn pulsing and no random component and emittance growth only in the vertical plane (solid line in Fig.~\ref{fig:10thsim}), the simulations predict qualitatively correctly an excitation amplitude dependent loss rate, but fail to predict the excitation amplitude dependent, continuous emittance growth observed in experiments.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}
Instead, the simulations yield a constant emittance after an initial increase during the first $10^4$~turns or first second. So obviously there is an essential ingredient missing. As it turns out, the constant emittance can be changed to an excitation amplitude dependent emittance growth by adding a random dipole noise component with uniform distribution representing the natural noise present in the LHC (dotted line in Fig.~\ref{fig:10thsim}). As no noise estimate is available at injection for the LHC, a first estimate is obtained by scaling the estimate for 6.5~TeV \cite{md1433_noise_top_energy,md_noise_bbLHC} by the magnetic rigidity to the injection energy of 450~GeV. This yields a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},
\end{equation}
Adding this random noise component, the simulations then yield correctly an excitation amplitude dependent emittance growth in the vertical plane after a first fast adjustment of the beam distribution visible as an increased initial value. Adding a random noise component has also the effect of increasing the loss rate. As the loss rates are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment is therefore further reduced, but a large difference still prevails as illustrated in Fig.~\ref{fig:10thexpsim} by means of a qualitative comparison of the loss rates obtained in experiments and simulations.
\begin{figure}[h]
		\centering
		\includegraphics[width=1.0\linewidth]{scale_amp_10v_ran_lbllong_sim.png}
	\caption{\label{fig:10thexpsim} Comparison of scaling of loss rates with the $10^{\mathrm{th}}$~turn pulsing excitation amplitude. The experimental results are shown in black (maximum excitation with 48~nrad) and blue (maximum excitation with 96~nrad). The excitation amplitude error is 50\% due to limited measurement accuracy and the error on the loss rate is the pure statistical error. The simulation results (2016 injection optics, standard errors, $(Q_x,Q_y)=(64.28,59.31)$, $10^{\mathrm{th}}$~turn pulsing in V plus random dipole noise in H+V of 6~nrad) are shown in red.}
\end{figure}

\subsubsection{8th turn pulsing\label{sec:simex8}}
\subsubsection{random excitation\label{sec:simexran}}

\subsubsection{7th turn pulsing\label{sec:simex7}}
The $7^{\mathrm{th}}$~turn pulsing pattern is the only pulsing pattern which has been tested during both experiments in 2016 and 2017 and was intended to serve as a proof for the reproducibility of the results. However, in 2017 the machine tune was changed from (64.28,59.31) to (62.27,60.295) in standard operation accompanied by a slight change in optics. The attempt to change back to the fractional tune of (.28,.31) as chosen in 2016 lead to large losses during the experiments in 2017 and therefore the 2017 working point of (62.27,60.295) was used during the experiment. This change in tune entailed a change in the driving resonances, from previously the $7Q_x$ resonances in 2016  (Fig.~\ref{fig:patternfma}) to the $7Q_x$, $7Q_y$ and $7Q_x + 7Q_y$ resonance in 2017 (Fig.~\ref{fig:7th2017fma}) and also a change in the affected amplitudes of the beam distribution.

@@

		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate.png}

@@

		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}

@@

	\caption{\label{fig:7th2017fma} FMA analysis for $7^{\mathrm{th}}$~turn pulsing based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition the 14th order $7Q_x+7Q_y$.}

@@

Consequently the effects on the beam also differ for both experiments (see Table~\ref{tab:7thexp} for a general overview and Fig.~\ref{} and Appendix~\ref{} for more details).
\begin{table}[h]
	\caption{\label{tab:7thexp}%
		Overview of the effect of $7^{\mathrm{th}}$~turn pulsing during the two resonant excitation experiments in 2016 and 2017~\cite{resexmd2016,resexmd2017}. The plane of the excitation is abbreviated with H for horizontal, V for vertical and H+V for horizontal and vertical at the same time.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Experiment 2017 & Experiment 2016  \\
			\colrule
			tune $(Q_x,Q_y)$ & (62.27,60.295) & (64.28,59.31)\\\hline
			excitation plane  & H,V and H+V&  H\\
			losses & similar loss rates for  & losses for pulsing \\
			& pulsing in H,V & in H \\
			&  and H+V & \\
			hor. emit. growth & no & yes, very small \\
			vert. emit. growth & yes, for pulsing in V & no\\
			&  and H+V & \\
		\end{tabular}
	\end{ruledtabular}
\end{table}
Simulations more or less fail completely (Fig.~\ref{fig:7thsim}). Higher amplitudes have been used for simulations to see noticable effect.

@@

		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_intensity.png}

@@

		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_rel_emit2.png}

@@

		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit1.png}

@@

		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit2.png}

@@

	\caption{\label{fig:7thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$ and 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates an excitation with only a random dipole noise component in H+V of 6~nrad, and the dotted and dashed line the results for $7^{\mathrm{th}}$~turn pulsing plus a random dipole noise component in H+V of 6~nrad.}

@@

\begin{itemize}
	\item Don't understand why there is emittance growth in H for 2016. There is no emittance growth in the experiment.
	\item repeat simulations for 2017 for all three planes and with random noise => check if there is only emittance growth in V \textcolor{red}{done}.
	\item experimentally: 2016 + 2017 losses in all cases, only emittance growth in V in 2017
\end{itemize}
\subsubsection{Effect of the transverse damper\label{sec:damp}}
Compare random and nth turn -> no damping for nth, only for random.

In general:
\begin{itemize}
	\item repeat simulations for each kth turn pattern with only b2 errors (\textcolor{green}{done})
	\item show results when adding random noise -> amplitude dependent emittance growth + losses
	\item FMA plot of 7th (2017) -> explain why in H and in V both losses, FMA plot of 10th (2016), FMA plot of 8th and random
	\item beam distribution change for 10th
\end{itemize}
\begin{itemize}
	\item 7th turn: small emittance but high losses
	\item 8th turn: almost no effect
	\item 10th turn: adjustment of distribution -> emittance growth + small losses
	\item random: constant filamentation -> emittance growth + later strong losses when beam hits aperature

\end{itemize}

\section{Summary\label{sec:sum}}

@@

\appendix
\section{FMA analysis for 10th turn pulsing and 2016 optics and tune}
\label{sec:fma:10}
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skh_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skv_dp0_ord10.png}
	\end{minipage}
	\caption{\label{fig:fma:10} FMA analysis for $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. There is no significant difference in the FMA analysis between pulsing only in H, only in V or in H+V.}
\end{figure}

\section{Results of $7^{\mathrm{th}}$~turn pulsing during 2016 and 2017 experiments}

>>>>

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total. Both schemes are illustrated in Fig.~\ref{fig:fill}.
\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment \textcolor{red}{(update picture)}
		\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
	\centering
	2017 experiment
	\includegraphics[width=1.0\linewidth]{bunchfilling.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom).}
\end{figure*}

@@

	\caption{\label{fig:patternfma} FMA analysis for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H further confirms that the $7Q_x$ resonance is the main resonance driven. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{app:sec:fma:10}, Fig.~\ref{app:fig:fma:10}).}

@@

\subsubsection{$10^{\mathrm{th}}$ turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing pattern was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item Increase of the loss rate with the excitation amplitude as shown in Figs.~\ref{fig:10thexp} and \ref{fig:10thexpsim}.
	\item Increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane shown in Fig.~\ref{fig:10thexp}.
	\item Change of the beam distribution due to the resonant excitation illustrated in Fig.~\ref{fig:10thexp} and \ref{fig:10thexpprof}.
	\item As loss rates, emittance growth and distribution stayed unchanged for the $4$~reference bunches, the above observations can be truly associated with the effect of the resonant excitation.	
\end{enumerate}
	\begin{minipage}[t]{1.0\linewidth}
		\includegraphics[width=0.49\linewidth]{2016_bunch_intensity_v10th_no_damper_avg.png}
		\includegraphics[width=1.0\linewidth]{2016_emith_avg_rel_v10th_no_damper.png}
		\includegraphics[width=1.0\linewidth]{2016_emitv_avg_rel_v10th_no_damper.png}
	\caption{\label{fig:10thexp} Relative intensity (top) and relative emittance (bottom) averaged over the bunches experiencing the same excitation amplitude during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}

@@

		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_50.png}

@@

		excited bunch,\\ $10^{\mathrm{th}}$~turn V		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_1300.png}

@@

	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. The black line in the plot of the residual is the current profile minus the profile before applying the excitation and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis see \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}

@@

@@

		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_intensity.png}

@@

		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2_rel.png}

@@

		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit1.png}

@@

		\includegraphics[width=1.0\linewidth]{2016injerra2b2uran1_2e-3_10thV_3_5um_emit2.png}

@@

	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}

@@

Besides the determination of the scaling of the loss rate and emittance growth with the excitation amplitude as required for the specification of the hollow electron lens, the most novel observation is the direct measurement of the change of the beam distribution, here with the Beam Synchrotron Radiation Telescope (BSRT)~\cite{bsrtprofinj}. As an example for the change in distribution, the vertical profile of a reference bunch and one bunch experiencing the maximum excitation of $5\cdot\Delta A=A_{\mathrm{max}}$ are compared in Fig.~\ref{fig:10thexpprof}. The distribution of the excited bunch clearly changes and becomes non-Gaussian, while the reference bunch stays unchanged.

Comparing these experimental results with the simulations including only the excitation through the $10^{\mathrm{th}}$~turn pulsing and no random component and emittance growth only in the vertical plane (solid line in Fig.~\ref{fig:10thsim}), the simulations predict qualitatively correctly an excitation amplitude dependent loss rate, but fail to predict the excitation amplitude dependent, continuous emittance growth observed in experiments. Instead, the simulations yield a constant emittance after an initial increase during the first $10^4$~turns or first second. So obviously there is an essential ingredient missing. As it turns out, the constant emittance can be changed to an excitation amplitude dependent emittance growth by adding a random dipole noise component with uniform distribution representing the natural noise present in the LHC (dotted line in Fig.~\ref{fig:10thsim}). As no noise estimate is available at injection for the LHC, a first estimate is obtained by scaling the estimate for 6.5~TeV \cite{md1433_noise_top_energy,md_noise_bbLHC} by the magnetic rigidity to the injection energy of 450~GeV. This yields a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},
\end{equation}
Adding this random noise component, the simulations then yield correctly an excitation amplitude dependent emittance growth in the vertical plane after a first fast adjustment of the beam distribution visible as an increased initial value. Adding a random noise component has also the effect of increasing the loss rate. As the loss rates are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment is therefore further reduced, but a large difference still prevails as illustrated in Fig.~\ref{fig:10thexploss} by means of a qualitative comparison of the loss rates obtained in experiments and simulations.
\begin{figure}[h]
		\centering
		\includegraphics[width=1.0\linewidth]{2016_scale_amp_10v_ran_lbllong_sim.png}
	\caption{\label{fig:10thexploss} Comparison of scaling of loss rates with excitation amplitude for $10^{\mathrm{th}}$~turn pulsing. The experimental results are shown in black (maximum excitation with 48~nrad) and blue (maximum excitation with 96~nrad). The excitation amplitude error is 50\% due to limited measurement accuracy and the error on the loss rate is the pure statistical error. The simulation results (2016 injection optics, standard errors, $(Q_x,Q_y)=(64.28,59.31)$, $10^{\mathrm{th}}$~turn pulsing in V plus random dipole noise in H+V of 6~nrad) are shown in red.}
\end{figure}

\subsubsection{$7^{\mathrm{th}}$ turn pulsing\label{sec:simex7}}
\begin{table}[h]
	\caption{\label{tab:7thexp}%
		Overview of the effect of $7^{\mathrm{th}}$~turn pulsing during the two resonant excitation experiments in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. The plane of the excitation is abbreviated with H for horizontal, V for vertical and H+V for horizontal and vertical at the same time.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Experiment 2017 & Experiment 2016  \\
			\colrule
			optics & 2017 injection optics & 2016 injection optics \\
			tune $(Q_x,Q_y)$ & (62.27,60.295) & (64.28,59.31)\\\hline
			excitation plane  & H,V and H+V&  H\\
			losses & similar loss rates for  & large losses for \\
			& pulsing in H,V &  pulsing in H \\
			&  and H+V & \\
			hor. emit. growth & no & yes, very small \\
			vert. emit. growth & yes, for pulsing in & no\\
			&  V and H+V & \\
		\end{tabular}
	\end{ruledtabular}
\end{table}
The $7^{\mathrm{th}}$~turn pulsing pattern has been tested during both experiments in 2016 and 2017 and was intended to serve as a proof for the reproducibility of the results. However, in 2017 the machine tune was changed from (64.28,59.31) to (62.27,60.295) in standard operation accompanied by a slight change in optics. The attempt to change back to the fractional tune of (.28,.31) as chosen in 2016 lead to large losses during the experiment in 2017 and therefore the 2017 working point of (62.27,60.295) was used instead. This change in tune entailed a change in the driving resonances, from previously the $7Q_x$ resonances in 2016 (Fig.~\ref{fig:patternfma}) to the $7Q_x$, $7Q_y$ and $7Q_x + 7Q_y$ resonance in 2017 (Fig.~\ref{fig:7th2017fma}).
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}
	\end{minipage}	
	\caption{\label{fig:7th2017fma} FMA analysis for $7^{\mathrm{th}}$~turn pulsing based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition the 14th order $7Q_x+7Q_y$.}
\end{figure}
Consequently, the effects on the beam also differ as can be illustrated by a comparison of the Distribution Tracking for both experiments~(Fig.~\ref{fig:7thsim}).
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_rel_emit2.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit2.png}
	\end{minipage}
	\caption{\label{fig:7thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$ and 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates an excitation with only a random dipole noise component in H+V of 6~nrad, and the dotted and dashed line the results for $7^{\mathrm{th}}$~turn pulsing plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}

\begin{figure}[h]
	%	\begin{minipage}[t]{0.49\linewidth}
	%		\centering
	%		2016 experiment
	%		\includegraphics[width=1.0\linewidth]{2016_bunch_intensity_h7th_no_damper_avg.png}
	%	\end{minipage}	
	%	\begin{minipage}[t]{0.49\linewidth}
	%		\centering
	%		2017 experiment
	%		\includegraphics[width=1.0\linewidth]{2017_bunch_intensity_hv7th_no_damper_avg.png}
	%	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		2016 experiment, $7^\mathrm{th}$ H	\includegraphics[width=1.0\linewidth]{2016_emith_avg_rel_h7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		2017 experiment, $7^\mathrm{th}$ H+V
		\includegraphics[width=1.0\linewidth]{2017_emith_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016_emitv_avg_rel_h7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:7thexp} Relative horizontal (top) and vertical (bottom) emittance obtained during the 2016 experiments (left) and the 2017 experiment (right) averaged over the bunches experiencing the same excitation amplitude.}
\end{figure}
Similar changes of the emittance are also experimentally observed as shown in Fig.~\ref{fig:7thexp}, where emittance growth was observed only in the horizontal plane for 2016 for pulsing in H, and only in the vertical plane for 2017 for pulsing in V and H+V (for pulsing in H no emittance growth is observed). Just as for the $10^{\mathrm{th}}$~turn pulsing (Fig.~\ref{fig:10thexp}), the emittance growth can be divided also for the $7^{\mathrm{th}}$~turn pulsing in a phase of fast adjustment of the beam distribution and the reach of a new equilibrium state (Fig.~\ref{fig:7thexp}). Also in this case, the distribution changes are also directly visible on the beam profiles measured with the BSRT (Fig.~\ref{fig:7thexpprof}).
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_hv7th_slot_2862.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ $7^{\mathrm{th}}$~turn h+V		\includegraphics[width=1.0\linewidth]{profile_hv7th_slot_2696.png}
	\end{minipage}
	\caption{\label{fig:7thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the $7^{\mathrm{th}}$~turn pulsing in H+V. The distribution changes in the reference bunch (left), here slot 2862, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure}

The loss rates measured during the experiment however differ from the simulations in terms of for which case the largest rate is observed as well as quantitatively for the specific cases. For example smaller losses are expected for the 2016 experiments compared to the 2017 experiments from simulations, while experimentally higher loss rates were measured in 2016 (Fig.~\ref{fig:7thexploss}).
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{2017_scale_amp_7h_lbllong.png}
	\caption{\label{fig:7thexploss} Comparison of scaling of loss rates with excitation amplitude for $7^{\mathrm{th}}$~turn pulsing in H and data obtained during the 2016 experiment (black and blue) and 2017 experiment (red and yellow). The excitation amplitude error is 50\% in 2016 and could be reduced to 15\% in 2017.}
\end{figure}

\subsubsection{$8^{\mathrm{th}}$ turn pulsing\label{sec:simex8}}
Pulsing patterns which have little effect on the beam core are particularly interesting for the hollow electron lens operation. With this motivation in mind the $8^{\mathrm{th}}$ turn pulsing pattern was studied, for which a small effect is expected in comparison to the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$ turn pulsing. In the Distribution Tracking, the excitation amplitudes had to be increased to 96~nrad to have a visible effect
\subsubsection{random excitation\label{sec:simexran}}

\subsubsection{Effect of the transverse damper\label{sec:damp}}
Compare random and nth turn -> no damping for nth, only for random.

\section{Summary\label{sec:sum}}
We now want to close the loop to the effects on the beam core for hollow electron lenses in HL-LHC. For DC operation, no influence on the beam core are are to be expected. To study the effect of pulsed operation beam experiments were conducted in 20216 and 2017 at injection. The experimental results showed that for the kick amplitudes expected for the e-lens, considerable effects are to be expected. The random pulsing mode is the most sensitive pulsing pattern and as it excites all beam frequencies results in emittance growth in the plane of excitation and losses once the beam size increased enough to hit the physical aperture. The resonant excitation or nth turn pulsing on the other hand features a rich non-linear dynamics as it excites only certain resonances and diversity in beam distribution and diffusion changes. This means, that each pulsing patterns results in different loss rates and emittance growth. Typical for this kind of excitation is also the adjustment of the beam distribution to a new equilibrium state.

@@

\appendix
\section{FMA analysis for 10th turn pulsing and 2016 optics and tune}
\label{app:sec:fma:10}
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skh_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skv_dp0_ord10.png}
	\end{minipage}
	\caption{\label{app:fig:fma:10} FMA analysis for $10^{\mathrm{th}}$~turn pulsing based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. There is no significant difference in the FMA analysis between pulsing only in H, only in V or in H+V.}
\end{figure}

\section{Results of $7^{\mathrm{th}}$~turn pulsing during 2016 and 2017 experiments (additional figures)}
\label{app:sec:simexp:7}
